The Impact of Accretion Disk Winds on the Optical Spectra of Cataclysmic Variables

### 2014-12-16 draft.tex
about to change conclusions on single peaked lines

It has been known for a long time that winds emanating from the
accretion disk are important in shaping the ultraviolet (UV) spectra
of high-state CVs (Heap 1978). The most spectacular evidence for such
outflows are the P-Cygni-like profiles seen in UV resonance lines such as
\civfull\ (see e.g. Cordova \& Mason
1982\nocite{cordova1982}). Considerable effort has been spent over the
years on understanding and modelling these UV features (e.g. Drew \&
Verbunt 1985\nocite{drewverbunt1985}; Mauche \& Raymond
1987\nocite{maucheraymond1987}; Drew 1987; Shlosman \& Vitello 1993; [hereafter
SV93]\nocite{SV93}; Knigge, Woods \& Drew 1995 [hereafter
KWD95]\nocite{KWD95}; Knigge \& Drew 1997\nocite{kd1997}; 
Knigge et al. 1997\nocite{knigge1997}; Long \&
Knigge 2002 [hereafter LK02]\nocite{LK02}, Noebauer et al. 2010;
Puebla et al. 2011\nocite{puebla2011}). The basic picture emerging from these efforts is
of a slowly accelerating, moderately collimated bipolar
outflow that carries away $\simeq 1\% - 10\%$ of the accreting
material. State-of-art-simulations of line formation in this type
of disk wind can produce UV line profiles that are remarkably similar
to observations.

Much less is known about the effect of these outflows on the optical
spectra of high-state CVs. These spectra are typically characterized
by H and He emissions superposed on a blue continuum. In many
cases, and particularly in the SW~Sex subclass of NLs
\citep{HSK86,DR95}, these lines are single-peaked. This is contrary to
theoretical expectations for lines formed in accretion disks, which
are predicted to be double-peaked \citep{smak1981, hornemarsh1986}. 
{\em Low-state} CVs (dwarf novae in quiescence) can, in fact,
exhibit such double-peaked lines \citep{marshhorne1990}. 
{\bf JM: I've included a reference to a paper on IP Peg here
which shows the transition from double peaked lines in 
quiescence and a contrast to outburst spectra. I'm not sure
of general trends in DNe. A lot seem to have weak emission cores
embedded in absorption.}

Murray \& Chiang (1996; hereafter MC96)\nocite{MC96} 
have shown that the presence of disk winds may
offer a natural explanation for the single-peaked optical emission lines in
high-state CVs, since they can strongly affect the radiative transfer
of line photons. Strong support for a significant wind contribution to the
optical emission lines comes from observations of eclipsing
systems. There, the single-peaked lines are often only weakly
eclipsed, and a significant fraction of the line flux remains visible
even near mid-eclipse (REF). This points to line formation in a spatially
extended region, such as a disk wind (see figure~\ref{novalikes}).
%%(Honeycutt, Schlegel, \& Kaitchuck 1986; Dhillon \& Rutten 1995). 
Further evidence for a wind contribution to the optical lines comes
from isolated observations of P-Cygni-like line profiles even in optical
lines, such as \ha\ and He \textsc{i} $\lambda5876$ \citep{RN98, kafka2004}.
{\bf JM: need to mention this later as pointed out by CK- both in discussion
and conclusion sections.}

@@

Attempts to fit the observed spectral energy distributions (SEDs) of
high-state CVs with such models have met with mixed success. In
particular, the SEDs predicted by most stellar/disk atmosphere models 
are too blue in the UV \citep{wade1988,long1991,long1994}(Other refs here?) and exhibit
stronger-than-observed Balmer jumps in absorption \cite{wade1984,haug1987,ladous1989b,knigge1998}(Other refs here?). One possible
explanation for these problems is that these models fail to capture
all of the relevant physics. Indeed, it has been argued that a
self-consistent treatment can 
produce better agreement with observational data (e.g. Shaviv et
al. 1991;  but also more recent work by Idan et al. 2010). 
\nocite{idanshaviv2010} \nocite{shaviv1991}. However, an alternative 
explanation, suggested by \cite{KLWB98}, is that recombination continuum
emission from the base of the disk wind might fill in the disk's
Balmer absorption edge and flatten the UV spectrum.

@@

The emissivity of these recombination 
features scales as $\rho^2$, meaning that they form almost entirely in the 
dense base of the wind, just above the accretion disk. Here, the
velocity field of the wind is still dominated by rotation, rather than
outflow, which accounts for the double-peaked shape of the lines. In
principle, lines formed in this region can still be single peaked,
since the existence of a poloidal velocity {\em gradient} changes the
local escape probabilities (MC96). However, as
demonstrated explicitly in section~\ref{modelb}, the line opacity in our
benchmark model is not high enough for this radiative transfer effect
to dominate the line shapes.

@@

Two other features are worth noting in the optical
spectrum. First, the collisionally excited Ca~{\sc ii} H\&K 
emission lines at 3969\AA\ and (especially) 3934\AA\ 
become quite prominent in our densest models. Second, our model predicts a detectable
He~\textsc{ii} recombination line at 3202\AA. This is the Helium
equivalent of Paschen~$\beta$ and should be expected in all systems that
feature a strong He~\textsc{ii}~4686\AA\ line (the Helium
equivalent of Paschen~$\alpha$). Although it has been observed (REF),
this line is somewhat unfamiliar observationally. 
This is because it lies bluewards of the atmospheric cut-off, but
also redwards of most ultraviolet spectra. 
% {\bf JM: we should we reference something here for the 3202 line. 
% I blieve CK found a paper which shows it in an IP? I'll try and find it.}

In the UV, the model still produces all the observed lines, 
and deep P-Cygni profiles are produced in the normal resonance lines,
as discussed in section ??. However, the UV spectra also
display what is perhaps the biggest problem with this revised model,
namely the strength of resonance line emission 
at low and intermediate inclinations.
In order to generate strong optical wind signatures, we have adopted wind
parameters that lead to very high densities at the base of the wind
($n_e\sim10^{13}-10^{14}$ in the main emission region). This produces
the desired optical recombination emission, but also increases the
role of collisional excitation in the formation of the UV resonance
lines. This explains the pronounced increase in the emission component 
of the C\textsc{iv} 1550\AA\ resonance line, for example, relative to
what was seen in the benchmark model (compare figures~\ref{spec} and
\ref{uvoptb}). The strength of this component in the revised model 
is probably somewhat too high to be consistent with UV observations 
of high-state CVs (REF).

@@

The wind now also has a clear effect on the continuum shape,
as shown by figure~\ref{modelb_escape}. The majority of the
escaping spectrum has now been reprocessed in some way by the wind,
either by electron scattering (the wind is now moderately Thompson-thick),
or by bound-free processes. This is demonstrated by the flatter spectral shape
and the slight Helum photoabsorption edges present in the optical spectrum.
In addition, figures~11 and 12 clearly demonstrate that the wind produces
a recombination continuum sufficient to completely fill in the Balmer jump
at high inclinations. \footnote{Note that the apparent absorption feature 
just redward of the Balmer jump in these models is artificial. It is
caused by residual line blanketing in the stellar atmospheres, which
our models cannot fill in since they employ a 20-level Hydrogen atom.}
This offers a potential solution to a long-standing problem in
understanding the spectral energy distribution of CVs, as
originally suggested by \cite{KLWB98}.

It has been noted by \cite{ladous1989, ladous1989b}
that the Balmer jump in high-state CVs would naturally weaken at
high inclinations due to limb darkening effects. 
Although we include a simple limb darkening law which affects 
the {\em total flux} seen at each incliantion, we cannot
include it as an effect on the {\em opacities} in our model.
As a result, the efficiency of filling in the Balmer jump
should really be judged at low and medium inclinations, 
where, although prominent, the recombination continuum does
not overcome the disk atmosphere absorption. 
In addition, this effect 
could mean that any model which successfully fills in the 
jump at low inclinations could lead to a Balmer jump 
in emission at high inclinations.
To understand this phenomenon, a fully self-consistent
radiative transfer calculation of both the disk atmosphere
and connected wind is required.

@@

% Although \ha\ is single-peaked in our model, the majority of the
% lines are double-peaked, particularly  
Figure~\ref{halpha} shows how the H$\alpha$ and 
% He~{\sc ii}~$\lambda 4686$ 
line changes with the kinematics of the wind 
for an inclination of $80^\circ$. The width and strength of He~{\sc ii} 
clearly depend on both $R_v$ and $\alpha$, but the overall line profile 
shape is always double-peaked. This contrasts with the behaviour of 
H$\alpha$, which becomes single-peaked in our densest models. The transition from
double-peaked to single-peaked occurs when the line becomes optically
thick. This reproduces the behaviour predicted by \cite{MC96}, who
showed that the poloidal velocity gradient in the wind can modify the
shape of optically thick emission lines even in the near-disk regime,
where the poloidal velocity itself is much smaller than the rotational
velocity. This has the effect of suppressing the line wings and
can therefore lead to the production of a single-peaked line profile
shape.

\begin{figure}
\includegraphics[width=0.5\textwidth]{figures/mc.eps}
\caption{
\ha\ line profiles, normalised to 1, plotted in veliocity space 
for three models with varying kinematic 
properties, computed at an inclination of $80^\circ$.
The benchmark model and the improved optical
model described in section 6 are labeled as A and B respectively,
and a third model (X) which has an increased acceleration length of 
$R_V = 283.8~R_{WD}$, and $\alpha=4$ is also shown. 
The $x$-axis limits correspond to the Keplerian velocity at 
$4R_{WD}$, the inner edge of the wind.
We observe a transition from double to singl-peaked lines
as the emission region becomes more optically thick.
% The parameter that has been altered from the benchmark model 
% is marked in each column and row. Models towards the bottom right
% are have more slowly accelerating winds and denser wind bases. 
% {\bf JM: Need to discuss the `single-peaked-ness' of line profiles
% in our telecon.}
}
\label{halpha}
\end{figure} %fullpage

@@

The similarity between the synthetic and observed spectra is
striking. In particular, the revised model produces strong emission in
all the Balmer lines, with line-to-continuum ratios comparable to
those seen in RW Tri. Moreover, the line-to-continuum contrast
increases during eclipse, as expected for emission produced in a disk
wind. This trend is in line with the observations of RW~Tri, and it
has also been seen in other NLs, including members of the SW~Sex class
(REF) 
% {\bf CK: I'm pretty sure that we know this about UX Uma, e.g. from
%   eclipse mapping studies, and there ought to be mid-eclipse spectra
%   of some additional system}
% The revised model also produes a single-peaked \ha\ line, in line with
% the observational data.

>>>>

It has been known for a long time that winds emanating from the
accretion disk are important in shaping the ultraviolet (UV) spectra
of high-state CVs (Heap 1978). The most spectacular evidence for such
outflows are the P-Cygni-like profiles seen in UV resonance lines such as
\civfull\ (see e.g. Cordova \& Mason
1982\nocite{cordova1982}). Considerable effort has been spent over the
years on understanding and modelling these UV features (e.g. Drew \&
Verbunt 1985\nocite{drewverbunt1985}; Mauche \& Raymond
1987\nocite{maucheraymond1987}; Drew 1987; Shlosman \& Vitello 1993; [hereafter
SV93]\nocite{SV93}; Knigge, Woods \& Drew 1995 [hereafter
KWD95]\nocite{KWD95}; Knigge \& Drew 1997\nocite{kd1997}; 
Knigge et al. 1997\nocite{knigge1997}; Long \&
Knigge 2002 [hereafter LK02]\nocite{LK02}, Noebauer et al. 2010\nocite{noebauer};
Puebla et al. 2011\nocite{puebla2011}). The basic picture emerging from these efforts is
of a slowly accelerating, moderately collimated bipolar
outflow that carries away $\simeq 1\% - 10\%$ of the accreting
material. State-of-art-simulations of line formation in this type
of disk wind can produce UV line profiles that are remarkably similar
to observations.

Much less is known about the effect of these outflows on the optical
spectra of high-state CVs. These spectra are typically characterized
by H and He emissions superposed on a blue continuum. In many
cases, and particularly in the SW~Sex subclass of NLs
\citep{HSK86,DR95}, these lines are single-peaked. This is contrary to
theoretical expectations for lines formed in accretion disks, which
are predicted to be double-peaked \citep{smak1981, hornemarsh1986}. 
{\em Low-state} CVs (dwarf novae in quiescence) can, in fact,
exhibit such double-peaked lines \citep{marshhorne1990}. 
% {\bf JM: I've included a reference to a paper on IP Peg here
% which shows the transition from double peaked lines in 
% quiescence and a contrast to outburst spectra. I'm not sure
% of general trends in DNe. A lot seem to have weak emission cores
% embedded in absorption.}

Murray \& Chiang (1996, 1997; hereafter MC96, MC97)\nocite{MC96, MC97} 
have shown that the presence of disk winds may
offer a natural explanation for the single-peaked optical emission lines in
high-state CVs, since they can strongly affect the radiative transfer
of line photons. Strong support for a significant wind contribution to the
optical emission lines comes from observations of eclipsing
systems. There, the single-peaked lines are often only weakly
eclipsed, and a significant fraction of the line flux remains visible
even near mid-eclipse (REF). This points to line formation in a spatially
extended region, such as a disk wind (see figure~\ref{novalikes}).
%%(Honeycutt, Schlegel, \& Kaitchuck 1986; Dhillon \& Rutten 1995). 
Further evidence for a wind contribution to the optical lines comes
from isolated observations of P-Cygni-like line profiles even in optical
lines, such as \ha\ and He \textsc{i} $\lambda5876$ \citep{RN98, kafka2004}.
{\bf JM: need to mention this later as pointed out by CK- both in discussion
and conclusion sections.}

@@

Attempts to fit the observed spectral energy distributions (SEDs) of
high-state CVs with such models have met with mixed success. In
particular, the SEDs predicted by most stellar/disk atmosphere models 
are too blue in the UV \citep{wade1988,long1991,long1994}(Other refs here?) and exhibit
stronger-than-observed Balmer jumps in absorption 
\citep{wade1984,haug1987,ladous1989b,knigge1998}. One possible
explanation for these problems is that these models fail to capture
all of the relevant physics. Indeed, it has been argued that a
self-consistent treatment can 
produce better agreement with observational data (e.g. Shaviv et
al. 1991;  but also more recent work by Idan et al. 2010).
\nocite{idanshaviv2010} \nocite{shaviv1991}
However, an alternative explanation, suggested by \cite{KLWB98}, 
is that recombination continuum emission from the base of the 
disk wind might fill in the disk's
Balmer absorption edge and flatten the UV spectrum.

@@

The emissivity of these recombination 
features scales as $\rho^2$, meaning that they form almost entirely in the 
dense base of the wind, just above the accretion disk. Here, the
velocity field of the wind is still dominated by rotation, rather than
outflow, which accounts for the double-peaked shape of the lines. In
principle, lines formed in this region can still be single peaked,
since the existence of a poloidal velocity {\em gradient} changes the
local escape probabilities (MC96, MC97). However, as
demonstrated explicitly in section~\ref{modelb}, the line opacity in our
benchmark model is not high enough for this radiative transfer effect
to dominate the line shapes.

@@

Two other features are worth noting in the optical
spectrum. First, the collisionally excited Ca~{\sc ii} H\&K 
emission lines at 3969\AA\ and (especially) 3934\AA\ 
become quite prominent in our densest models. Second, our model predicts a detectable
He~\textsc{ii} recombination line at 3202\AA. This is the Helium
equivalent of Paschen~$\beta$ and should be expected in all systems that
feature a strong He~\textsc{ii}~4686\AA\ line (the Helium
equivalent of Paschen~$\alpha$). 
%%Although it has been observed (REF),
This line is somewhat unfamiliar observationally, because it 
lies bluewards of the atmospheric cut-off, but
also redwards of most ultraviolet spectra. 
% {\bf JM: we should we reference something here for the 3202 line. 
% I blieve CK found a paper which shows it in an IP? I'll try and find it.}

In the UV, the model still produces all the observed lines, 
and deep P-Cygni profiles are produced in the normal resonance lines,
as discussed in section ??. However, the UV spectra also
display what is perhaps the biggest problem with this revised model,
namely the strength of resonance line emission 
at low and intermediate inclinations.
In order to generate strong optical wind signatures, we have adopted wind
parameters that lead to very high densities at the base of the wind
($n_e\sim10^{13}-10^{14}$ in the main emission region). This produces
the desired optical recombination emission, but also increases the
role of collisional excitation in the formation of the UV resonance
lines. This explains the pronounced increase in the emission component 
of the C\textsc{iv} 1550\AA\ resonance line, for example, relative to
what was seen in the benchmark model (compare figures~\ref{spec} and
\ref{uvoptb}). The strength of this component in the revised model 
is probably somewhat too high to be consistent with UV observations 
of high-state CVs (see e.g. Long et al. 1991, 1994; Noebauer et al. 2010).
\nocite{long1991,long1994, noebauer}

@@

The wind now also has a clear effect on the continuum shape,
as shown by figure~\ref{modelb_escape}. The majority of the
escaping spectrum has now been reprocessed in some way by the wind,
either by electron scattering (the wind is now moderately Thompson-thick),
or by bound-free processes. This is demonstrated by the flatter spectral shape
and the slight Helum photoabsorption edges present in the optical spectrum.
In addition, figures~11 and 12 clearly demonstrate that the wind produces
a recombination continuum sufficient to completely fill in the Balmer jump
at high inclinations.\footnote{Note that the apparent absorption feature 
just redward of the Balmer jump in these models is artificial. It is
caused by residual line blanketing in the stellar atmospheres, which
our models cannot fill in since they employ a 20-level Hydrogen atom.}
This offers a potential solution to a long-standing problem in
understanding the spectral energy distribution of CVs, as
originally suggested by \cite{KLWB98}.

It has been noted by \cite{ladous1989, ladous1989b}
that the Balmer jump in high-state CVs would naturally weaken at
high inclinations due to limb darkening effects. 
Although we include a simple limb darkening law which affects 
the seen at each inclination, we cannot
include it as a {\em frequency dependent} opacity in our model.
As a result, the efficiency of filling in the Balmer jump
should really be judged at low and medium inclinations, 
where, although prominent, the recombination continuum does
not overcome the disk atmosphere absorption. 
In addition, this effect 
could mean that any model which successfully fills in the 
jump at low inclinations could lead to a Balmer jump 
in emission at high inclinations.
To understand this phenomenon, a fully self-consistent
radiative transfer calculation of both the disk atmosphere
and connected wind is required.

@@

% Although \ha\ is single-peaked in our model, the majority of the
% lines are double-peaked, particularly  
Figure~\ref{halpha} shows how the H$\alpha$
% He~{\sc ii}~$\lambda 4686$ 
line changes with the kinematics of the wind
for an inclination of $80^\circ$. 
% The width and strength of He~{\sc ii} 
% clearly depend on both $R_v$ and $\alpha$, but the overall line profile 
% shape is always double-peaked. This contrasts with the behaviour of 
The line becomes single-peaked in our densest models. 
The transition from double-peaked to single-peaked occurs when the 
line becomes optically thick. This reproduces the behaviour predicted by MC96, who
showed that the poloidal velocity gradient in the wind can modify the
shape of optically thick emission lines even in the near-disk regime,
where the poloidal velocity itself is much smaller than the rotational
velocity. This has the effect of suppressing the line wings and
can therefore lead to the production of a single-peaked line profile
shape.

\begin{figure}
\includegraphics[width=0.5\textwidth]{figures/mc.eps}
\caption{
\ha\ line profiles, normalised to 1, plotted in veliocity space 
for three models with varying kinematic 
properties, computed at an inclination of $80^\circ$.
The benchmark model and the improved optical
model described in section 6 are labeled as A and B respectively,
and a third model (X) which has an increased acceleration length of 
$R_V = 283.8~R_{WD}$, and $\alpha=4$ is also shown. 
The $x$-axis limits correspond to the Keplerian velocity at 
$4R_{WD}$, the inner edge of the wind.
We observe a transition from double to single-peaked lines
as the emission region becomes more optically thick.
% The parameter that has been altered from the benchmark model 
% is marked in each column and row. Models towards the bottom right
% are have more slowly accelerating winds and denser wind bases. 
% {\bf JM: Need to discuss the `single-peaked-ness' of line profiles
% in our telecon.}
}
\label{halpha}
\end{figure} %fullpage

@@

The similarity between the synthetic and observed spectra is
striking. In particular, the revised model produces strong emission in
all the Balmer lines, with line-to-continuum ratios comparable to
those seen in RW Tri. Moreover, the line-to-continuum contrast
increases during eclipse, as expected for emission produced in a disk
wind. This trend is in line with the observations of RW~Tri, and it
has also been seen in other NLs, including members of the SW~Sex class
\citep{neustroev2011}.
% {\bf CK: I'm pretty sure that we know this about UX Uma, e.g. from
%   eclipse mapping studies, and there ought to be mid-eclipse spectra
%   of some additional system}
% The revised model also produes a single-peaked \ha\ line, in line with
% the observational data.
